Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models

### 2020-09-23 analysis/paper/paper.Rmd
a few more small revisions, based on an internal review. Removed comments that were no longer relevant.

---
title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

```{r packages}
library(survey)
library(tidyverse)
library(QRFpaper)
library(janitor)
library(knitr)
library(pander)
library(magrittr)
library(sf)
library(english)
library(kableExtra)

theme_set(theme_bw())

@@

# run this if running chunks directly
# setwd('analysis/paper')
```

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020]. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020]. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such data are unavailable in most watersheds [@Cramer2009], and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. Bioenergetics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the 200 - 600 m reach scale [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Wall2016; @Carmichael2020]. In addition, those authors did not take the step of scaling the capacity predictions at the reach scale to entire watersheds. In contrast, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr at the watershed scale, using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking for the watershed scale in the Pacific Northwest.

<!-- How's that transition sentence at the end of the paragraph above? -->

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho salmon (*O. kisutch*) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O. nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook salmon (*O. tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have used other modeling approaches to elicit fish-habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O. clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm..

@@

* Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
* Elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
* Extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
* Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations. Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish-habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other "H's."

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watersheds within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, single-pass electrofishing, and snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

Abundance estimates at all sites were then translated into linear (parr/m) fish densities which were paired with the associated CHaMP habitat data. For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity. After removing duplicate sites, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)). We did explore using areal fish densities (parr/m$^2$) as the response, but found very similar results so in the interest of brevity we only present results based on linear fish densities.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].  We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.

@@

All code and data for the analyses presented here can be found in a [GitHub repository](https://github.com/KevinSee/QRFpaper).

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

@@

There are two aspects that make this approach "data hungry", meaning a large dataset is needed to fit a QRF model like this. First, random forest models generally require more data that parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables. Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework. For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th. Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumeably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.

@@

<!-- Do we need any more in this section? I commented out the paragraph below... -->

<!-- While the Columbia River basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging. On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.  Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….  -->

<!-- Add here why GAA are messy. And then anything else which may be a consideration for using the extrapolation model moving forward.  -->

@@

<!-- Discussion of improving habitat data to possibly avoid extrapolation. Collect cheaper/better spatially continuous habitat data. -->

<!-- I have no idea if this is what you are looking for so feel free to delete if I am way off base.  -->

@@

<!-- The beginning of the following paragraph may need some word-smithing -->

@@

Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Office of Species Conservation through contract BOR002 16. Fish sampling work in the Lemhi River was also funded through the Idaho Office of Species Conservation through the Pacific Coast Salmon Recovery Fund. Special thanks to staff from the following agencies for providing data to calculate Chinook salmon parr abundance and density estimates: Columbia River Inter-Tribal Fish Commission, Oregon Department of Fish and Wildlife, and U.S. Fish and Wildlife Service. The models in this study were improved by conversations with Eric Buhle.

@@

```{r wtsd-map, out.width = "100%", fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
knitr::include_graphics('../figures/ChampWtds.png')
```

>>>>

---
title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile random forest
highlights: |
  These are the highlights. 
---

@@

```{r packages}
# run this if running chunks directly
# setwd('analysis/paper')

# what packages are necessary
library(survey)
library(tidyverse)
library(QRFpaper)
library(janitor)
library(knitr)
library(pander)
library(magrittr)
library(sf)
library(english)
library(kableExtra)
library(ggpubr)

theme_set(theme_pubr())

@@

if(doc.type %in% c("docx", "html")) {
  theme_set(theme_pubr(base_family = "Times New Roman"))
}

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020]. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020]. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of abundance for adults and juveniles, with variation in the number of adults. Such data are unavailable in most watersheds [@Cramer2009], and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. Bioenergetics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the 200 - 600 m reach scale [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Wall2016; @Carmichael2020]. In addition, those authors did not take the step of scaling the capacity predictions at the reach scale to entire watersheds. In contrast, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr at the watershed scale, using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking at the watershed scale in the Pacific Northwest.

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985 and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho salmon (*O. kisutch*) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O. nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook salmon (*O. tshawytscha*) in the Yukon River and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have used other modeling approaches to elicit fish-habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O. clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; @Compton2012]. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm.

@@

* identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
* elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
* predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
* extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
* validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations. Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish-habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated, and the magnitude of required change can be placed in context with the other "H's."

@@

CHaMP sites are 200 m to 600 m reaches in wadeable streams across select watersheds within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, single-pass electrofishing, and snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew/watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

Abundance estimates at all sites were then translated into linear (parr/m) fish densities which were paired with the associated CHaMP habitat data. For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity. After removing duplicate sites, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)). We did explore using areal fish densities (parr/m$^2$) as the response but found very similar results so in the interest of brevity we only present results based on linear fish densities.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].  We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.

@@

All code and data for the analyses presented here can be found in a GitHub repository ([(https://github.com/KevinSee/QRFpaper)](https://github.com/KevinSee/QRFpaper)).

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation that may be necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stages and/or specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

@@

There are two aspects that make this approach "data hungry", meaning a large dataset is needed to fit a QRF model like this. First, random forest models generally require more data than parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables. Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework. For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th. Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.

@@

@@

@@

@@

Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Governor's Office of Species Conservation through contract BOR002 16. Fish sampling work in the Lemhi River was also funded through the Idaho Governor's Office of Species Conservation through the Pacific Coast Salmon Recovery Fund. Special thanks to staff from the following agencies for providing data to calculate Chinook salmon parr abundance and density estimates: Columbia River Inter-Tribal Fish Commission, Oregon Department of Fish and Wildlife, and U.S. Fish and Wildlife Service. The models in this study were improved by conversations with Eric Buhle.

@@

```{r wtsd-map, out.width = "90%", fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
knitr::include_graphics('../figures/ChampWtds.png')
```
